---
title: "ggvis: Next generation Grammar of Graphics"
author: "Abhijit Dasgupta <br/>@webbedfeet <br/>abhijit@datacommunitydc.org"
date: "March 12, 2015"
output: ioslides_presentation
runtime: shiny
---
<style>
body{background-color:white;}
article{color:black;}
</style>
```{r preamble, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(ggvis)
```

## What is ggvis?

+ A data visualization package in R
+ Builds graphics declaratively using layers in the spirit of [ggplot2](http://ggplot2.org)
+ Produces graphics rendered in a web browser using [*Vega*](https://github.com/trifacta/vega/), 
a visualization grammar built on top of D3
+ Produces interactive graphics leveraging the power of Shiny


## ggplot2 vs ggvis

```{r, fig.height=4.5}
ggplot(mtcars, aes(x=wt, y=mpg))+
  geom_point()
```

## ggplot2 vs ggvis

```{r, fig.height=3.5}
mtcars %>% 
  ggvis(x= ~wt, y= ~mpg) %>% 
  layer_points()
```

## ggplot2 vs ggvis

```{r, fig.height=3.5}
mtcars %>% 
  ggvis(x= ~wt, y= ~mpg) %>% 
  layer_points() %>% layer_smooths()
```

## ggplot2 vs ggvis

ggplot2   |  ggvis
----------|---------------
geom      | layer (names end with "s")
stat      | compute
aes()     | props()

<p>

- ggvis uses a function interface, so chaining is done using `%>%` rather than `+`
- ggvis uses `x = ~wt` instead of `x = wt` (will come back to this)
- Using `ggvis()` without any layers is akin to `qplot`

# ggvis semantics

## Basic template
```{r, results='hide',eval=FALSE}
<data>  %>% 
  ggvis(~<x property>,~<y property>, 
        fill = ~<fill property>, ...) %>% 
  layer_<marks>()`

```


`<marks>` can be
 `arcs`, `bars`, `densities`, `freqpolys`,
`histograms`, 
`images`, 
`lines`, 
`paths`, 
`model_predictions`, `points`, `rects`, `ribbons`, `smooths`, `text`

## Components of a ggvis visualization

```{r, eval=FALSE}
faithful %>% 
    ggvis(~waiting, ~eruptions, fill := "red") %>% 
    layer_points() %>% 
    add_axis("y", title = "Duration of eruption (m)", 
             values = c(2, 3, 4, 5), subdivide = 9) %>% 
    add_axis("x", title = "Time since previous eruption (m)")

```
<p>

+ Data
+ A coordinate system
+ Marks
+ Corresponding properties

## Components of a ggvis visualization

```{r, echo=FALSE}
faithful %>% 
    ggvis(~waiting, ~eruptions, fill := "red") %>% 
    layer_points() %>% 
    add_axis("y", title = "Duration of eruption (m)", 
             values = c(2, 3, 4, 5), subdivide = 9) %>% 
    add_axis("x", title = "Time since previous eruption (m)")

```

# Some new semantics

## `=` and `:=`

+ `=` *maps* a property to a data value or a set of data values
+ `:=` *sets* a property to a specific value

```{r, fig.height=3}
pressure %>% ggvis(~temperature, ~pressure, fill = "red") %>% 
  layer_points()
```


## `=` and `:=` 

+ `=` *maps* a property to a data value or a set of data values 
+ `:=` *sets* a property to a specific value



```{r, fig.height=3}
pressure %>% ggvis(~temperature, ~pressure, fill := "red") %>% 
  layer_points()
```

## `=` and `:=`

See [http://ggvis.rstudio.com/properties-scales.html](http://ggvis.rstudio.com/properties-scales.html) for more details.

# More new semantics

## Objects in ggvis

ggvis code accepts three types of objects: 

+ A string of letters, without quotes, refers to an *object* name, and will be searched in the current environment
+ A tilde `~` at the start of the string will be treated as a *variable* name, and will be searched in the current data frame
+ A string in quotes will be treated as a *raw value*. 

## Objects in ggvis

```{r, fig.height=3}
shade <- 'red'
mtcars %>% ggvis(~disp, ~mpg) %>% 
  layer_points(fill= ~factor(cyl), size :=300, stroke := shade, 
               strokeWidth :=6) %>% add_legend('fill',title='Cylinders')
```

# Examples

## Interplay of dplyr and ggvis
```{r, fig.height=3.5, fig.width=6, warning=FALSE, message=FALSE}
mtcars %>% group_by(am) %>% 
  ggvis(~mpg, ~hp) %>% layer_smooths(stroke = ~factor(am)) %>% 
    layer_points(fill = ~factor(am))

```

## Interplay of dplyr and ggvis
```{r, fig.height=3.5, fig.width=6, warning=FALSE, message=FALSE}
mtcars %>% group_by(am) %>% 
  ggvis(~mpg, ~hp) %>% layer_smooths(stroke = ~factor(am)) %>% 
    layer_points(fill = ~factor(am)) %>% 
  scale_nominal('fill', range=c("green","orange")) %>% 
  scale_nominal('stroke', range=c('green','orange'))
```

## Using derived statistics ([link](http://rpackages.ianhowson.com/cran/ggvis/man/dplyr-ggvis.html))
```{r, fig.height=3}
base <- mtcars %>% ggvis(~mpg, ~cyl) %>% layer_points()
base %>% group_by(cyl) %>% summarise(mpg = mean(mpg)) %>%
  layer_points(fill := "red", size := 100)
```

## Interactive visualization

```{r, fig.height=2, warning=FALSE, message=FALSE, echo=FALSE}
mtcars %>% ggvis(~wt, ~mpg) %>% layer_points() %>% 
 layer_smooths(se=T, span=input_slider(0.1,0.9,value=0.75,step=0.05,
                                       label='span'))
```

## Interactive visualization
```{r, fig.height=2, warning=FALSE, message=FALSE, eval=FALSE}
mtcars %>% ggvis(~wt, ~mpg) %>% layer_points() %>% 
 layer_smooths(se=T,
   span=input_slider(0.1,0.9,value=0.75,step=0.05,label='span'))
```

## Tooltips
```{r,echo=FALSE} 
mtcars %>% mutate(name=row.names(mtcars)) %>% 
  ggvis(~wt, ~mpg, key := ~name) %>% layer_points(size :=100) %>% 
  add_tooltip(function(df) df$name)
```

## Tooltips
```{r,eval=F, eval=FALSE}
mtcars %>% mutate(name=row.names(mtcars)) %>% 
  ggvis(~wt, ~mpg, key := ~name) %>% layer_points() %>% 
  add_tooltip(function(df) df$name)

```


## Summarizing models
```{r, fig.height=3, warning=FALSE, message=FALSE, error=FALSE}
mtcars %>% ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_model_predictions(model = "lm", stroke:='black') %>%
  layer_model_predictions(model = "loess", stroke := "red") 
```

## Resources

+ [ggvis.rstudio.com](http://ggvis.rstudio.com)
+ [Tutorial](http://ggvis.rstudio.com/ggvis-basics.html)
+ [Additional examples](http://rpackages.ianhowson.com/cran/ggvis/)
+ [Interactivity](http://ggvis.rstudio.com/interactivity.html)
+ [Customization](http://ggvis.rstudio.com/properties-scales.html)
+ A ggplot and ggvis comparison by [Jim Hester](http://jimhester.github.io/ggplot2ToGgvis/)

## Maps (from Bob Rudis' [site](http://rud.is/b/2014/12/29/making-static-interactive-maps-with-ggvis-using-ggvis-maps-wshiny/))

![][id]
```{r, echo=FALSE, eval=FALSE, message=FALSE, error=FALSE, warning=FALSE}
library(ggvis)
library(ggplot2)
library(rgdal)
library(rgeos)
library(magrittr)
library(dplyr)
library(RColorBrewer)
library(data.table)
library(maptools)

us <- readOGR("data/us.geojson", "OGRGeoJSON")
us <- us[!us$STATEFP %in% c("02", "15", "72"),]

us_aea <- spTransform(us, CRS("+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs"))

map <- ggplot2::fortify(us_aea, region="GEOID")

droughts <- read.csv("data/dm_export_county_20141223.csv")
droughts$id <- sprintf("%05d", as.numeric(as.character(droughts$FIPS)))
droughts$total <- with(droughts, (D0+D1+D2+D3+D4)/5)

map_d <- merge(map, droughts, all.x=TRUE)

ramp <- colorRampPalette(c("white", brewer.pal(n=9, name="YlOrRd")), space="Lab")

map_d$fill_col <- as.character(cut(map_d$total, seq(0,100,10), include.lowest=TRUE, labels=ramp(10)))
map_d$fill_col <- ifelse(is.na(map_d$fill_col), "#FFFFFF", map_d$fill_col)

# drought_values <- function(x) {
#   if(is.null(x) | !(x$id %in% droughts$id)) return(NULL)
#   y <- droughts %>% filter(id==x$id) %>% select(1,3,4,6:10)
#   sprintf("<table width='100%%'>%s</table>",
#           paste0("<tr><td style='text-align:left'>", names(y),
#          ":</td><td style='text-align:right'>", format(y), collapse="</td></tr>"))
# }

map_d %>%
  group_by(group, id) %>%
  ggvis(~long, ~lat) %>%
  layer_paths(fill:=~fill_col, strokeOpacity := 0.5, strokeWidth := 0.25) %>%
  #add_tooltip(drought_values, "hover") %>%
  #Çhide_legend("fill") %>%
  hide_axis("x") %>% hide_axis("y")# %>%
  #set_options(width=900, height=600, keep_aspect=TRUE)

```

[id]: data/plot_231142944.svg
